---
title: "Research Log Project Machine Learning"
subtitle: "Differentiating between benign and malignant samples of breast masses"
output: pdf_document
header-includes:
   - \pagenumbering{gobble}
   - \usepackage{longtable}
   - \usepackage{hyperref}
---


<!-- (Front page) -->
\vspace{350pt}

\hfill \textbf{Student}: Vincent Talen  

\hfill \textbf{Student number}: 389015  

\hfill \textbf{Class}: BFV3

\hfill \textbf{Study}: Bio-Informatics

\hfill \textbf{Institute}: Institute for Life Science & Technology

\hfill \textbf{Teachers}: Dave Langers (LADR) and Bart Barnard (BABA)

\hfill \textbf{Date}: `r Sys.Date()`

\newpage
<!-- (Table of contents) -->
\setcounter{secnumdepth}{2}
\tableofcontents
\pagenumbering{arabic}


\newpage
<!-- (Setting up R) -->
# Setting up R
For the data analysis and further actions a couple of libraries are needed, they are loaded in here.
A few other options/settings are also configured here.
```{r setup}
# Set code chunks visibility in pdf output to true
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.align = "center")

# Create vector with all packages that will be used
packages <- c("readr", "ggplot2", "gridExtra", "scales", "ggpubr", "reshape2")
# Load each package in the vector with lapply
invisible(lapply(packages, library, character.only = TRUE))
# Drop the packages variable from memory since it will not be used again
remove(packages)
```

\newpage
<!-- (Data Set) -->
# Data Set
## Origin of the data
Van onderzoekje jwz.

## Structure of data set
To easily annotate figures and to better understand and distinguish the data a codebook was manually created. This codebook will now be imported and the explanations of the multiple columns shown.
```{r import codebook}
# Import codebook
codebook <- read_delim("data/codebook.txt", delim = "|", show_col_types = FALSE)
codebook[,c(1,4)]
```

## Loading in the data
```{r load data from file}
# Load in data from file
data <- read_csv("data/wdbc.data", col_names = codebook[[1]], show_col_types = FALSE)
head(data)
```

```{r data amount}
# Print the amount of samples and columns
cat("Amount of samples:", dim(data)[1], "\tColumns in dataframe", dim(data)[2])
```

<!-- (References) -->
\newpage
\begin{thebibliography}{9}

\bibitem{street93}
W.N. Street, W.H. Wolberg and O.L. Mangasarian. (1993), \textit{Nuclear feature extraction for breast tumor diagnosis.}, 1993 International Symposium on Electronic Imaging: Science and Technology, volume 1905, pages 861-870, \url{https://minds.wisconsin.edu/bitstream/handle/1793/59692/TR1131.pdf} (accessed Sep 16, 2022).

\end{thebibliography}