---
output:
  pdf_document:
    includes:
      in_header: ../report_subfiles/import.tex
      before_body: ../report_subfiles/before_body.tex

knit: (function(inputFile, encoding) {
  rmarkdown::render( inputFile, encoding = encoding, output_dir = here::here() )})

bibliography: ../report_subfiles/references.bib
nocite: '@*'
urlcolor: blue
---

```{r setup, include=FALSE}
# Set code cunk options
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.align = "center")

# Set directory of this Rmd file to project directory
knitr::opts_knit$set(root.dir = here::here())
options("knitr.graphics.rel_path" = FALSE)
```

\newpage
<!-- (Introduction) -->
# Introduction
Diagnosing if breast masses/lumps are benign or malignant, and thus if they are breast cancer, was done by performing full biopsies, which are invasive surgical procedures. An alternative method is to take fine needle aspirate (FNA) fluid samples and then have the acting physician try to determine the diagnosis by looking with a microscope at the cell nuclei. The problem with this, and why they thus mainly performed full biopsies, was that this gave mixed results because it was highly subjective and depended a lot on the skill of the physician. But because diagnosing the breast masses by taking FNAs and examining them is much less invasive than the full biopsies it was desired to make this process faster and improve the correctness and objectivity of the diagnosis process. To achieve this goal microscopic images were digitized of the FNA samples and processed into features, machine learning techniques were then used to create a diagnostic model.

## Objective
Can breast mass malignancy be assessed with >95% accuracy, using visual features of nuclei boundaries computed from fine needle aspirate images?

## Theory
??


\newpage
<!-- (Materials and Methods) -->
# Materials and Methods
Link to GitHub Repository of Part 1: Analysis:  
[https://github.com/Vincent-Talen/Project-Machine-Learning-Part1_Analysis](https://github.com/Vincent-Talen/Project-Machine-Learning-Part1_Analysis)  

Link to GitHub Repository of Part 2: Java Wrapper:  
[https://github.com/Vincent-Talen/Project-Machine-Learning-Part2_Java-Wrapper](https://github.com/Vincent-Talen/Project-Machine-Learning-Part2_Java-Wrapper)  

## Materials
??

## Existing Methods
??

## Developed Methods
??


\newpage
<!-- (Results) -->
# Results
## Data
```{r data_distribution, out.width = "75%", fig.cap = "Barplot showing distribution of diagnosis classification labels"}
knitr::include_graphics(here::here("src/rmd/research_log_figures/data-distribution-1.pdf"))
```


\newpage
<!-- (Discussion and Conclusion) -->
# Discussion and Conclusion
## Discussion
??

## General conclusion and perspective
??


<!-- (After-Body) -->
```{r after body, child = 'src/report_subfiles/after_body.Rmd'}
```