---
output:
  pdf_document:
    includes:
      in_header: ../report_subfiles/import.tex
      before_body: ../report_subfiles/before_body.tex

knit: (function(inputFile, encoding) {
  rmarkdown::render( inputFile, encoding = encoding, output_dir = here::here() )})

bibliography: ../report_subfiles/references.bib
nocite: '@*'
---

```{r, include=FALSE}
# Copyright (c) 2022 Vincent Talen.
# Licensed under GPLv3. See LICENSE file.
```

```{r setup, include=FALSE}
# Set code cunk options
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.align = "center")

# Set directory of this Rmd file to project directory
knitr::opts_knit$set(root.dir = here::here())
options("knitr.graphics.rel_path" = FALSE)
```

\newpage
<!-- (Introduction) -->
# Introduction
Diagnosing if breast masses/lumps are benign or malignant, and thus if they are breast cancer, was done by performing full biopsies, which are invasive surgical procedures. An alternative method is to take fine needle aspirate (FNA) fluid samples and then have the acting physician try to determine the diagnosis by looking with a microscope at the cell nuclei. The problem with this, and why they thus mainly performed full biopsies, was that this gave mixed results because it was highly subjective and depended a lot on the skill of the physician. But because diagnosing the breast masses by taking FNAs and examining them is much less invasive than the full biopsies it was desired to make this process faster and improve the correctness and objectivity of the diagnosis process. To achieve this goal microscopic images were digitized of the FNA samples and processed into features, machine learning techniques were then used to create a diagnostic model.

## Objective
Can breast mass malignancy be assessed with >95% accuracy, using visual features of nuclei boundaries computed from fine needle aspirate images?

## Theory
??


\newpage
<!-- (Materials and Methods) -->
# Materials and Methods
Link to GitHub Repository of Part 1: Analysis:  
[https://github.com/Vincent-Talen/Project-Machine-Learning-Part1_Analysis](https://github.com/Vincent-Talen/Project-Machine-Learning-Part1_Analysis)  

Link to GitHub Repository of Part 2: Java Wrapper:  
[https://github.com/Vincent-Talen/Project-Machine-Learning-Part2_Java-Wrapper](https://github.com/Vincent-Talen/Project-Machine-Learning-Part2_Java-Wrapper)  

## Materials
??

## Existing Methods
??

## Developed Methods
??


\newpage
<!-- (Results) -->
# Results
## Data
```{r data_distribution, out.width = "75%", fig.cap = "Barplot showing distribution of diagnosis classification labels"}
knitr::include_graphics(here::here("output/figures/data-distribution-1.pdf"))
```


\newpage
<!-- (Discussion and Conclusion) -->
# Discussion and Conclusion
## Discussion
??

## General conclusion and perspective
??


\newpage
<!-- (Project proposal for Minor ) -->
# Project proposal for Minor 
The proposal is for a project for the minor High- Throughput High Performance Biocomputing (HTHPB) of the Bachelor Bio-Informatics at the Hanze University of Applied Sciences Groningen. 

The project that was done now was to create a model with machine learning algorithms that can accurately predict breast mass malignancy using the features of nucleus boundaries. These features however are collected by a tedious process, the acting physician has to create an image with a video camera that looks through a microscope at a preparation of a taken fine needle aspirate sample. Then on a computer they have to manually roughly trace the outlines of each nucleus' boundary in the program Xcyt, after which the program computes the features that are eventually used.

The goal of the project that is proposed is to improve the process of gathering and computing these features by creating a new program that can go through a lot of digital images of preparations without the intervention of humans. It should thus be able to identify the outlines of the nucleus boundaries with a neural network algorithm. 

This newly created algorithm can then be connected with the model, and potentially the java wrapper, of this project, together creating a single tool that takes the digital image taken and in one go returns the prediction of the breast mass malignancy. The speed of diagnosing would thus be greatly improved and the acting physicians would have to spend less valuable time on these tedious tasks that can be automated.


<!-- (After-Body) -->
```{r after body, child = 'src/report_subfiles/after_body.Rmd'}
```